---
title: "Linear Modeling in R - Predicting NYC condominium sale prices"
author: "Bena"
date: "`r Sys.Date()`"
output: html_document
---

## Background

In this project we'll use condominium sales data from all 5 boroughs of New York City to determine how well the size of a condominium (measured in gross square feet) explains/predicts sale price for each individual borough & across New York City as a whole. This property sales data is publicly available here (https://data.world/dataquest/nyc-property-sales-data) and covers sales records from a twelve-month period (from November 1, 2018, through October 31, 2019).
The goal of the project is to apply our skills particularly linear regression models (aka bivariate linear regression), evaluate summaries of regression statistics and assess the model quality.

## Understanding the data

```{r}
library(tidyverse)
library(ggplot2)
library(stats)
```

Importing the data

```{r}
NYC_sales <- suppressMessages(read_csv("NYC_property_sales.csv"))
head(NYC_sales)
```

```{r}
glimpse(NYC_sales)
```

Finding out the most popular building classifications 

```{r}
sort(table(NYC_sales$building_class_at_time_of_sale))
```
"R4" was the most popular building classification. It's a condominium;residential unit in elevator building.
Refer here (https://www.nyc.gov/assets/finance/jump/hlpbldgcode.html) for the building classifications as represented by the codes above.
We'll focus our analysis on this type of property only, for now.

```{r}
NYC_condos <- NYC_sales %>%
               filter(building_class_at_time_of_sale == "R4")
NYC_condos
```

## Explore Bivariate Relationships with Scatterplots

```{r}
ggplot(data = NYC_condos,
       aes(x = gross_square_feet, y = sale_price)) +
  geom_point(aes(color = borough), alpha = 0.5) + 
  scale_y_continuous(labels = scales::comma) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Plot of sale price against size",
       x = "Size (gross square feet)",
       y = "Sale Price (USD)")
```

Zooming in on a smaller subset of the data

```{r}
ggplot(data = NYC_condos,
       aes(x = gross_square_feet, y = sale_price)) + 
  geom_point(aes(color = borough), alpha = 0.5) + 
  scale_y_continuous(labels = scales::comma, limits = c(0,50000000)) + 
  xlim(0,5000) +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Plot of sale price against size",
       x = "Size (gross square feet)",
       y = "Sale Price (USD)")
```

```{r}
sort(table(NYC_condos$borough))
```

```{r}
ggplot(data = NYC_condos,
       aes(x = gross_square_feet, y = sale_price)) + 
  geom_point(alpha = 0.5) + 
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ borough, ncol = 2, scales = "free") +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Plot of sale price against size",
       x = "Size (gross square feet)",
       y = "Sale Price (USD)")
```

## Outliers and Data Integrity Issues

We begin with sorting the data from the most to the least expensive

```{r}
NYC_condos <- NYC_condos %>%
                arrange(-sale_price)
head(NYC_condos, 10)
```

```{r}
tail(NYC_condos, 10)
```

The first two data points, valued above $200 million, seem to be out of the sales price range of the rest of the data.
Upon investigation, we found that the property valued at $239,958,219 is indeed one of the most expensive homes sold in the U.S. The collection of properties has their own webpage and this specific property is listed therein. Thus, we shall retain it.
The second property valued at $200,000,000 seems to be a purchase of several units. We shall remove this from the dataset since the analysis is limited to transactions of single units for now, with an interest of predicting sales prices. 
It will also be good to find out if there are any other multiple sales recorded as one transaction, since these misrepresent the prices.

```{r}
NYC_condos2 <- NYC_condos %>%
                filter(sale_price != 200000000)
head(NYC_condos2)
```

In general sales of multiple units will have a recording of the same date and price.

```{r}
multi_unit_sales <- NYC_condos2 %>% 
                      group_by(sale_price, sale_date) %>% 
                      filter(n() >= 3) %>% 
                      arrange(desc(sale_price))
multi_unit_sales
```

Sites like this (https://streeteasy.com/) also helped us determine whether the prices were within range, in addition to searching street address on Google.

```{r}
brooklyn <- NYC_condos2 %>%
              filter(borough == "Brooklyn") %>%
              arrange(desc(sale_price))
```

what's the mean for gross_square_feet & sale_price for each borough?

```{r}
mean_gross_sqft <- NYC_condos2 %>%
                             group_by(borough) %>%
                             summarize(mean_gross_square_feet = mean(gross_square_feet))
                            
mean_gross_sqft
```

```{r}
mean_price <- NYC_condos2 %>%
               group_by(borough) %>%
               summarize(mean_sale_price = mean(sale_price))
                            
mean_price
```

investigate the entries in excel****
These could also be investigated further to half the price, in case we'll need to improve the accuracy of our model further (we shall revisit of need be)

```{r}
multi_unit_sales2 <- NYC_condos2 %>% 
                      group_by(sale_price, sale_date) %>% 
                      filter(n() == 2) %>% 
                      arrange(desc(sale_price))
```

Removing the multi-sale units i.e. for 3 or more units. The code below drops all observations in _NYC_condos2_ that have a match in _multi_unit_sales_.

```{r}
NYC_condos2 <- NYC_condos2 %>%
                 anti_join(multi_unit_sales)
NYC_condos2

#Alternatively
#NYC_condos2 <- NYC_condos2 %>%
#                group_by(sale_price, sale_date) %>%
#                filter(n() <= 2) %>%
#                ungroup()
#NYC_condos2
```

Mean for gross_square_feet & sale_price for each borough after cleaning

```{r}
mean_gross_sqft2 <- NYC_condos2 %>%
                             group_by(borough) %>%
                             summarize(mean_gross_square_feet = mean(gross_square_feet))
                            
mean_gross_sqft2
```


```{r}
mean_price2 <- NYC_condos2 %>%
               group_by(borough) %>%
               summarize(mean_sale_price = mean(sale_price))
                            
mean_price2
```

## Linear Regression Model for Boroughs in New York City Combined

We will generate bivariate linear regression models for all New York City boroughs combined, before & after removing sales of multiple units for comparison.

```{r}
NYC_condos_lm <- lm(sale_price ~ gross_square_feet, data = NYC_condos)
summary(NYC_condos_lm)
NYC_condos2_lm <- lm(sale_price ~ gross_square_feet, data = NYC_condos2)
summary(NYC_condos2_lm)
```

The hypotheses state that:<br/>
   >H0: there is a relationship between the size of a condominium (`gross_square_feet`) and the price (`sale_price`)<br/>
   >H1: there is no relationship between the size of a condominium (`gross_square_feet`) and the price (`sale_price`)<br/>
   
We can declare there is a relationship between condominium size and price when the slope is sufficiently far from zero.
The slope for the original NYC_condos data is 1192.72 while that for NYC_condos2 data is 4462. Both are far from 0 hence there's a relationship between condominium size and price.

For each model, the t-statistic was high enough, and the p-value was low enough, to declare that there is, in fact, a relationship between `gross_square_feet` and `sale_price`. The t-statistic for the cleaned dataset (`NYC_condos2`) was nearly double that of the original dataset (`NYC_condos`) at 113.04 versus 61.39. 

In each case the p-value was well below the 0.05 cutoff for significance meaning that it is extremely unlikely that the relationship between condominium size and sale price is due to random chance. 

```{r}
4.462e+03 == 4462
```

```{r}
confint(NYC_condos_lm)
confint(NYC_condos2_lm)
```

The confidence interval for the slope for original NYC_condos dataset is [1154.636, 1230.802].
The confidence interval for the slope for NYC_condos2 dataset is [4384.254, 4538.999].
This difference can likely be attributed to the removal of many multi-million dollar sale records for smaller units which impacted price predictions in the original dataset.

The measure for *lack of fit*, or residual standard error (RSE) for original NYC_condos dataset is 4,745,000 while that for the NYC_condos2 dataset is lower i.e., 2,945,000. Also note that the NYC_condos2 dataset is less than the original NYC_condos dataset by 150 observations (the one transaction for $200,000,000 plus those in multi-unit-sales).

```{r}
RSE_NYC_condos <- sigma(NYC_condos_lm)
RSE_NYC_condos
RSE_NYC_condos2 <- sigma(NYC_condos2_lm)
RSE_NYC_condos2
```

The R-squared, or the proportion of the variability in `sale_price` that can be explained by `gross_square_feet` in original NYC_condos dataset is 0.3177 while that for the NYC_condos2 dataset is 0.6166 (almost double). Implying that the linear model did a better job in predicting property sale price after the records for multiple sales were removed. The model explains about 61.66% of variability in sale prices.

Regenerating the scatterplot faceted by borough using the cleaned dataset.

```{r}
ggplot(data = NYC_condos2,
       aes(x = gross_square_feet, y = sale_price)) + 
  geom_point(alpha = 0.5) + 
  scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ borough, ncol = 2, scales = "free") +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Plot of sale price against size",
       x = "Size (gross square feet)",
       y = "Sale Price (USD)")
```

Above is the updated scatterplot that uses the cleaned `NYC_condos` data. For the Brooklyn borough we are better able to see the spread of the data and how the trend line fits the data because we removed the \$30 million outliers. 
The same is true for the Manhattan borough because the $200 million multi-unit sale was removed.

## Linear Regression Models for each Borough - Coefficient Estimates

We'll use the broom package & workflow to generate and compare linear models for each borough.
The general workflow using broom and tidyverse tools to generate many models involves 4 steps:<br/>
  1.Nest a dataframe by a categorical variable with the `nest()` function from the `tidyr` package - we will nest by `borough`.<br/>
  2.Fit models to nested dataframes with the `map()` function from the `purrr` package.<br/>
  3.Apply the `broom` functions `tidy()`, `augment()`, and/or `glance()` using each nested model - we'll work with `tidy()` first.<br/>
  4.Return a tidy dataframe with the `unnest()` function - this allows us to see the results.<br/>
  
```{r}
library(broom)
```

```{r}
NYC_nested <- NYC_condos2 %>%    
                group_by(borough) %>%
                nest()
NYC_nested
```

We can extract and inspect the values of any nested dataframe. Below is a look at the first few rows for Queens

```{r}
head(NYC_nested$data[[4]])
```

Fit linear models to each borough, individually

```{r}
NYC_coefficients <- NYC_condos2 %>%    
                      group_by(borough) %>%
                      nest() %>%
                      mutate(linear_model = map(.x=data,
                                                .f = ~lm(sale_price ~ gross_square_feet, data = .)))
NYC_coefficients
```

We can view the linear modeling results for any one of the nested objects using the `summary()` function. 
Below are the linear regression statistics for Manhattan

```{r}
summary(NYC_coefficients$linear_model[[1]])
```

A quick look at the R-squared value for the Queens linear model indicates that `gross_square_feet` looks to be a good single predictor of `sale_price`. 63.37% of variability in Sale price in Manhattan can be explained by the model.
Transforming these linear model summary statistics into a tidy format.

```{r}
NYC_coefficients <- NYC_condos2 %>%    
                      group_by(borough) %>%
                      nest() %>%
                      mutate(linear_model = map(.x=data,
                                                .f = ~lm(sale_price ~ gross_square_feet, data = .))) %>%
                      mutate(tidy_coefficients = map(.x = linear_model,
                                                     .f  = tidy, conf.int = TRUE))
NYC_coefficients
```

Unnest to a tidy dataframe of coefficient estimates.

```{r}
NYC_tidy <- NYC_coefficients %>%
             select(borough, tidy_coefficients) %>%
             unnest(cols = tidy_coefficients)
NYC_tidy
```

Filter to return the slope estimate only

```{r}
NYC_slope <- NYC_tidy %>%
              filter(term == "gross_square_feet") %>%
              arrange(estimate)
NYC_slope
```

We've arranged the results in ascending order by the slope estimate. 
For each of the 5 boroughs, the t-statistic and p-value indicate that there is a relationship between `sale_price` and `gross_square_feet`. In Staten Island, an increase in square footage by one unit is estimated to increase the sale price by about \$288 on average. In contrast, an increase in total square footage by one unit in Manhattan is estimated to result in an increase in sale price of about \$4,728 on average.
We can infer that, in general, a property in Staten Island will cost less per square foot than in Manhattan i.e., an increase in size will cost less per square foot in Staten Island than in Manhattan.

## Linear Regression Models for each Borough - Regression Summary Statistics

```{r}
NYC_stats <- NYC_condos2 %>%    
                      group_by(borough) %>%
                      nest() %>%
                      mutate(linear_model = map(.x=data,
                                                .f = ~lm(sale_price ~ gross_square_feet, data = .))) %>%
                      mutate(tidy_summary_stats = map(.x = linear_model,
                                                      .f = glance))
NYC_stats
```

We unnest the results of the glance function

```{r}
NYC_tidy2 <- NYC_stats %>%
              select(borough, tidy_summary_stats) %>%
              unnest(cols = tidy_summary_stats) %>%
              arrange(r.squared)
NYC_tidy2
```

```{r}
NYC_nested2 <- NYC_condos2 %>%
                group_by(borough) %>%
                nest() %>%
                mutate(linear_model = map(.x=data,
                                          .f = ~lm(sale_price ~ gross_square_feet, data = .))) %>%
                mutate(tidy_coefficients = map(.x = linear_model,
                                               .f  = tidy, conf.int = TRUE)) %>%
                mutate(tidy_summary_stats = map(.x = linear_model,
                                                .f = glance)) %>%
                mutate(data_augmented = map2(.x = linear_model,
                                            .y = data,
                                            .f = augment))
NYC_nested2
```

```{r}
NYC_augmented <- NYC_nested2 %>%
                  select(borough, data_augmented) %>%
                  unnest(cols = data_augmented)
NYC_augmented
```

An interpretation of the fitted & residual columns generated by the augment function is that the model had estimated that the first property with a sale price of \$239,958,219 would cost \$105,624,640.6 but it was sold for way more than that. Thus the difference between the actual and estimated price was \$134,333,578.38. Also referred to as the residual.

## Conclusion

Our analysis showed that, in general, the `gross_square_feet` variable is useful for explaining, or estimating, `sale_price` for property sales in New York City. 
We observed that removing multi-unit sales from the dataset increased model accuracy, from 31.77% to 61.66%. However, there is still room for improvement. 
With linear models generated for New York City as a whole, and with linear models generated for each borough individually, we observed in all cases that the t-statistic was high enough, and the p-value was low enough, to declare that there is a relationship between `gross_square_feet` and `sale_price`. In the near future, we could use this dataset to perform multivariate linear regression i.e., predicting sale price using `gross_square_feet` and `land_square_feet` where the data is available.

For the linear models that we generated for each individual borough, we observed a wide range in slope estimates. The slope estimate for Manhattan was much higher than the estimate for any of the other boroughs. We did not remove the record-setting \$240 million property sale from the dataset, but future analysis should investigate the impacts that this single listing has on modeling results. 

Finally, regression summary statistics indicate that `gross_square_feet` is a better single predictor of `sale_price` in some boroughs versus others. For instance, the R-squared value was estimated at approximately 0.34 and 0.48 in Queens and Staten Island compared to an estimate of 0.63 in Manhattan. These differences in R-squared correspond with the scatterplots generated for each borough; the strength of sale prices versus gross square feet was higher, and the dispersion (spread), was lower for Manhattan compared to Queens where the relationship was moderate because the data was more spread out. 












