---
title: "Linear Modeling in R: Predicting Property Sale Prices"
author: "Bena"
date: "`r Sys.Date()`"
output: html_document
---

## Background

In this project we'll use condominium sales data from all 5 boroughs of New York City to determine how well the size of a condominium (measured in gross square feet) explains/predicts sale price for each individual borough & across New York City as a whole. This property sales data is [publicly available] (https://www.nyc.gov/site/finance/property/property-rolling-sales-data.page) and contains sales records from a twelve-month period (February,2023 - January,2024)
The goal of the project is to apply our skills particularly linear regression models (aka bivariate linear regression), evaluate summaries of regression statistics and assess the model quality.

## Understanding the data

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
```

Importing the data

```{r}
bronx <- read_excel("rollingsales_bronx.xlsx", skip = 4)
brooklyn <- read_excel("rollingsales_brooklyn.xlsx", skip = 4)
manhattan <- read_excel("rollingsales_manhattan.xlsx", skip = 4)
queens <- read_excel("rollingsales_queens.xlsx", skip = 4)
statenisland <- read_excel("rollingsales_statenisland.xlsx", skip = 4)
```

Bind the dataframes

```{r}
NYC_property_sales <- bind_rows(bronx, brooklyn, manhattan, queens, statenisland)
glimpse(NYC_property_sales)
```
Remove the dataframes from each borough from memory

```{r}
rm(bronx, brooklyn, manhattan, queens, statenisland)
```

Change column names so they're easy to work with

```{r}
library(janitor)
NYC_property_sales <- NYC_property_sales %>%
                          janitor::clean_names()
```

Alternatively

```{r}
#colnames(NYC_property_sales) %<>% str_replace_all("\\s", "_") %>% tolower()
```

The borough column in each dataframe is a numeric code which we'll replace with the respective names i.e.,
1 - manhattan, 2 - bronx, 3 - brooklyn, 4 - queens, 5 - statenisland
Replace borough code with borough name

```{r}
NYC_property_sales <- NYC_property_sales %>%
                       mutate(borough = case_when(
                        borough == 1 ~ "manhattan",
                        borough == 2 ~ "bronx",
                        borough == 3 ~ "brooklyn",
                        borough == 4 ~ "queens",
                        borough == 5 ~ "statenisland"
                       ))
```

```{r}
unique(NYC_property_sales$borough)
```
Convert CAPITALIZED columns to Title Case

```{r}
NYC_property_sales <- NYC_property_sales %>%
                       mutate(neighborhood = str_to_title(neighborhood),
                              building_class_category = str_to_title(building_class_category),
                              address = str_to_title(address))
```

Retain only distinct observations & drop columns that won't be used i.e. all values in `easement` are NA  

```{r}
unique(NYC_property_sales$easement)
```
```{r}
NYC_property_sales <- NYC_property_sales %>%
                       select(-easement) %>%
                       distinct()
```

Filtering data

```{r}
NYC_property_sales2 <- NYC_property_sales %>%
                        filter(gross_square_feet > 0) %>%
                        #to remove property that changed hands among family members, we assumed USD 10,000 is the threshold
                        filter(sale_price >= 10000) %>%  
                        drop_na(c(gross_square_feet,sale_price))

glimpse(NYC_property_sales2)
```

A glimpse of the data reveals we're left with 21,089 observations.
Arrange observations alphabetically by borough and neighborhood.

```{r}
NYC_property_sales2 <- NYC_property_sales2 %>%
                         arrange(borough, neighborhood)
```

Save the results to a CSV file

```{r}
write.csv(NYC_property_sales2, "NYC_property_sales2.csv")
#Commented out after running code to avoid accidentally rewriting
```

Load the csv file into R

```{r}
property_sales <- suppressMessages(read_csv("NYC_property_sales2.csv"))
```

## Explore Bivariate Relationships with Scatterplots
To examine the following characteristics: direction, linearity, and strength.

Filter property_sales to include only the most popular building classification (refer to link for building types https://www.nyc.gov/assets/finance/jump/hlpbldgcode.html) 

```{r}
sort(table(property_sales$building_class_at_time_of_sale))
```

A1 looks like the most popular. They are one family dwellings designed as two stories - SM OR MID.

```{r}
property_sales2 <- property_sales %>%
                    filter(building_class_at_time_of_sale == "A1")
```

Generate scatterplots

```{r}
ggplot(data = property_sales2,
       aes(x = gross_square_feet, y = sale_price)) + 
  geom_point(aes(color = borough), alpha = 0.5) + 
  scale_y_continuous(labels = scales::comma) + 
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Plot of sale price against size",
       x = "Size (gross square feet)",
       y = "Sale Price (USD)")
```

Zooming in on a smaller subset of the data

```{r}
ggplot(data = property_sales2,
       aes(x = gross_square_feet, y = sale_price)) + 
  geom_point(aes(color = borough), alpha = 0.5) + 
  scale_y_continuous(labels = scales::comma, limits = c(0,5000000)) + 
  xlim(0,5000) +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Plot of sale price against size",
       x = "Size (gross square feet)",
       y = "Sale Price (USD)")
```

We observe the same trend that sale price increases with increase in size.
The pattern is somewhat linear, but there is a fair amount of spread, or dispersion, that becomes more pronounced with an increase in condominium size.

Using facet wrap

```{r}
ggplot(data = property_sales2,
       aes(x = gross_square_feet, y = sale_price)) + 
  geom_point(alpha = 0.5) + 
   scale_y_continuous(labels = scales::comma) +
  facet_wrap(~ borough, ncol = 2, scales = "free") +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(title = "Plot of sale price against size",
       x = "Size (gross square feet)",
       y = "Sale Price (USD)")
```

Looking at the plot above, we see that, in general, larger condominiums are associated with a higher sale price in each borough. The data follows a somewhat linear pattern in each plot. But the spread is difficult to see with the Manhattan scatterplot, potentially because there are only 2 data points.
There is no obvious curvature with the shape of the data, for any borough. 
The strength of the bivariate relationship is weak for Bronx, moderate for Queens & Brooklyn. Staten Island borough looks like it has a stronger relationship between sale price and size. 

## Outliers and Data Integrity Issues

We begin with sorting the data from the most to the least expensive

```{r}
property_sales2 <- property_sales2 %>%
                     arrange(-sale_price)
head(property_sales2, 10)
```

```{r}
tail(property_sales2, 10)
```

The sale prices seem to be within range.<br/>
Checking if there are sales for multiple units recorded as one

```{r}
multi_unit_sales <- property_sales2 %>% 
                      group_by(sale_price, sale_date) %>% 
                      filter(n() >= 3) %>% 
                      arrange(desc(sale_price))
```

Looks like there are none.

## Linear Regression Model for Boroughs in New York City Combined

We'll now check the accuracy of our linear regression model

```{r}
property_sales2_lm <- lm(sale_price ~ gross_square_feet, data = property_sales2)
summary(property_sales2_lm)
```

The hypotheses state that:<br/>
   >H0: there is a relationship between the size of a condominium (`gross_square_feet`) and the price (`sale_price`)<br/>
   >H1: there is no relationship between the size of a condominium (`gross_square_feet`) and the price (`sale_price`)<br/>

We can declare there is a relationship between condominium size and price when the slope is sufficiently far from zero.
The slope in this case is 437.05 which is well above 0.
The t-statistic was high enough, and the p-value was low enough, to declare that there is, in fact, a relationship between `gross_square_feet` and `sale_price`.
The p-value is well below the 0.001 cutoff for significance meaning that it is extremely unlikely that the relationship between condominium size and sale price is due to random chance. 

```{r}
2e-16 == 0.0000000000000002
```
```{r}
confint(property_sales2_lm)
```

The confidence interval for the slope is [411.6165, 462.4876]
The measure for *lack of fit*, or residual standard error (RSE) is 483,400.
The R-squared, or the proportion of the variability in `sale_price` that can be explained by `gross_square_feet` is 0.2463. Meaning that only 24.6% of the variability in sale price can be explained by the regression model. This is low.

```{r}
RSE <- sigma(property_sales2_lm)
RSE
```

If we would have cleaned the data to remove some datapoints, then we would compare the above statistics for the datasets then regenerated the scatterplot with the cleaned data.

## Linear Regression Models for each Borough - Coefficient Estimates

We'll use the broom package & workflow to generate and compare linear models for each borough.
The general workflow using broom and tidyverse tools to generate many models involves 4 steps:<br/>
  1.Nest a dataframe by a categorical variable with the `nest()` function from the `tidyr` package - we will nest by `borough`.<br/>
  2.Fit models to nested dataframes with the `map()` function from the `purrr` package.<br/>
  3.Apply the `broom` functions `tidy()`, `augment()`, and/or `glance()` using each nested model - we'll work with `tidy()` first.<br/>
  4.Return a tidy dataframe with the `unnest()` function - this allows us to see the results.<br/>

```{r}
library(broom)
```

```{r}
property_sales2_nested <- property_sales2 %>%
                            group_by(borough) %>%
                            nest()
property_sales2_nested
```

We can extract and inspect the values of any nested dataframe. Below is a look at the first few rows for Queens.

```{r}
head(property_sales2_nested$data[[2]])
```

Fit linear models to each borough, individually

```{r}
property_sales2_coefficients <- property_sales2 %>%
                            group_by(borough) %>%
                            nest() %>%
                            mutate(linear_model = map(.x=data,
                                                      .f = ~lm(sale_price~gross_square_feet, data = .)))
property_sales2_coefficients
```

We can view the linear modeling results for any one of the nested objects using the `summary()` function. 
Below are the linear regression statistics for Queens.

```{r}
summary(property_sales2_coefficients$linear_model[[2]])
```
A quick look at the R-squared value for the Queens linear model indicates that `gross_square_feet` looks to be a good single predictor of `sale_price`. 39.9% of variability in Sale price in Queens can be explained by the model.
Transforming these linear model summary statistics into a tidy format

```{r}
property_sales2_coefficients <- property_sales2 %>%
                            group_by(borough) %>%
                            nest() %>%
                            mutate(linear_model = map(.x=data,
                                                      .f = ~lm(sale_price~gross_square_feet, data = .))) %>%
                            mutate(tidy_coefficients = map(.x=linear_model,
                                                           .f=tidy,
                                                           conf.int=TRUE))
                            
property_sales2_coefficients
```
Unnest to a tidy dataframe of coefficient estimates.

```{r}
property_sales2_tidy <- property_sales2_coefficients %>%
                          select(borough, tidy_coefficients) %>%
                          unnest(cols = tidy_coefficients)
property_sales2_tidy
```

Filter to return the slope estimate only

```{r}
property_sales2_slope <- property_sales2_tidy %>%
                           filter(term == "gross_square_feet") %>%
                           arrange(estimate)
property_sales2_slope
```

We've arranged the results in ascending order by the slope estimate. 
For each of the 5 boroughs, the t-statistic and p-value indicate that there is a relationship between `sale_price` and `gross_square_feet`. In Staten Island, an increase in square footage by one unit is estimated to increase the sale price by about \$262, on average. In contrast, an increase in total square footage by one unit is estimated to result in an increase in sale price of about \$4,728??, on average.
We can infer that, in general, a property in Bronx will cost less per square foot than in Staten Island i.e., an increase in size will cost less per square foot in Bronx than in Staten Island.

## Linear Regression Models for each Borough - Regression Summary Statistics

```{r}
property_sales2_stats <- property_sales2 %>%
                           group_by(borough) %>%
                           nest() %>%
                           mutate(linear_model = map(.x=data,
                                                      .f = ~lm(sale_price~gross_square_feet, data = .))) %>%
                           mutate(tidy_summary_stats = map(.x=linear_model,
                                                           .f=glance))
property_sales2_stats
```
We unnest the results of the glance function

```{r}
property_sales2_tidy <- property_sales2_stats %>%
                          select(borough, tidy_summary_stats) %>%
                          unnest(cols = tidy_summary_stats) %>%
                          arrange(r.squared)
property_sales2_tidy
```

Our analysis showed that, in general, the `gross_square_feet` variable is useful for explaining, or estimating, `sale_price` for property sales in New York City. 
We observed that removing multi-unit sales from the dataset increased model accuracy. 
With linear models generated for New York City as a whole, and with linear models generated for each borough individually, we observed in all cases that the t-statistic was high enough, and the p-value was low enough, to declare that there is a relationship between `gross_square_feet` and `sale_price`.

For the linear models that we generated for each individual borough, we observed a wide range in slope estimates. The slope estimate for Manhattan was much higher than the estimate for any of the other boroughs. We did not remove the record-setting \$240 million property sale from the dataset, but future analysis should investigate the impacts that this single listing has on modeling results. 

Finally, regression summary statistics indicate that `gross_square_feet` is a better single predictor of `sale_price` in some boroughs versus others. For instance, the R-squared value was estimated at approximately 0.43 in Staten Island & 0.39 in Queens, compared to an estimate of only 0.07 in Bronx. These differences in R-squared correspond with the scatterplots generated for each borough; the strength of sale prices versus gross square feet was higher, and the dispersion (spread), was lower for Staten Island and Queens as compared to Bronx where the relationship was noticeably weaker because the data was more spread out. Manhattan had only 2 data points perhaps because the *A1* property type are not popular there. The graph & summary statistics generated therefore could not be used to make inferences for the data.


```{r}

```












